How cryptic is cryptic diversity? Machine learning approaches to plastral variation in \textit{Emys marmorata}.

### 2013-07-15 documents/turtle_paper.Rnw
more draft work

\documentclass[12pt]{article}
\usepackage{amsmath, amsthm}
\usepackage{graphicx, microtype, hyperref, authblk}
\usepackage{rotating, longtable, caption, subcaption, multirow}
\usepackage[sort&compress]{natbib}
%\usepackage{fullpage}

@@

\frenchspacing

@@

\title{How cryptic is cryptic diversity? Machine learning approaches to plastral variation in \textit{Emys marmorata}.}
\author[1]{Peter D Smits \thanks{psmits@uchicago.edu}}
\author[2]{Kenneth D Angielczyk \thanks{kangielczyk@fieldmuseum.org}}
\author[3]{James F Parham \thanks{jparham@fullerton.edu}}
\affil[1]{Committee on Evolution Biology, University of Chicago}
\affil[2]{Department of Geology, Field Museum of Natural History}
\affil[3]{Department of Geological Sciences, California State University -- Fullerton}

\begin{document}

\maketitle

\linenumbers
\modulolinenumbers[2]

\begin{abstract}
  % 200 words

\end{abstract}

\section{Introduction}

% cryptic diversity
%   most species are still deliminated solely based on morphology
%   paleo problem
Cryptic diversity is when taxa were only first deliminated via molecular means and were not or cannot deliminated via morphological identification CITATION. The discovery of this previously unknown diversity has
% all the work done on species delimitation strictly from molecular data
%

@@

% GP superimposition and PCA
``Half'' plastra landmark configurations were superimposed using generalized Procrustes analysis \citep{Dryden1998a} after which, the principal components of shape were calculated. This was done using the \texttt{shapes} package for R \citep{2013, Dryden2013}.

@@

% model training
%   tuning parameters
%     grid search
%       what are the tuning parameters for each model?
%     best AUC ROC
For both types of supervised learning models, tuning parameters were estimated using 10 rounds of 10-fold cross-validation (CV) across a grid search of all tuning parameter combinations. Optimal tuning parameter values were selected based on area under the receiver operating characteristic curve (AUC ROC). Multiclass AUC ROC was estimated using the all-against one strategy derived by \citet{Hand2001} in implemented in PROC PACKGE.

@@

\subsubsection{Supervised learning}
For all classification schemes, the optimal random forest model based on recursive feature selection by maximizing AUC ROC of the model was one with many features (Fig. \ref{fig:roc})
% need to discuss the final models in both cases.
% figure
%   ROC model selection (see talk)
%   generalize densities
%     facet: multinomial, rf (see talk)

@@

This pattern of misclassification may be caused by the differences in mean shape between each of the different classes (Fig. \ref{fig:mean_shape}). The mean shape of the northern clade is the most similar to the mean shape of the entire dataset (Fig. ), which may mean that specimens that are closer to the mean shape will be systematically misclassified as the northern clade.

@@

% methodological concerns
%   compromise in the supervised learning models
%     fair because as much variation as ``necessary'' is used
%   unsupervised learning
%     model based clustering
%     nonparametric bayesian approach
%     anything about semisupervised clustering? probably not
Ultimately, it would be optimal to not require such explicit classification hypotheses, especially when concerned about possible cryptic variation in extinct taxa. The method employed in this study, PAM, is rather simple and not model based. Comparison is facilitated by comparison of a summary statistic and bootstrap confidence intervals. A more useful and future avenue would be employing various model based clustering approaches CITATIONS. In this manner, a series of candidate models can be compared via model comparison methods, such as AIC or Bayes factors CITATION, in order to asses the best clustering solution. Of particular note are nonparametric Bayesian approaches to model based clustering CITATIONS. This approach uses a class of flexible priors to allow for the most optimal clustering solution to be decided from the data. Currently, there exists a nonparametric Bayesian clustering method and further development of this approach may prove extremely fruitful for better deliminating taxa solely from morphology.

@@

\section*{Acknowledgements}
PDS would like to thank David Bapst, Michael Foote, Benjamin Frable, and Dallas Krentzel for useful discussion which enhanced the quality of this study.

>>>>

\documentclass[12pt,letterpaper]{article}
\usepackage{amsmath, amsthm}
\usepackage{graphicx, microtype, hyperref, authblk}
\usepackage{rotating, longtable, caption, subcaption, multirow}
\usepackage[sort&compress]{natbib}
\usepackage{fullpage}

@@

\frenchspacing
\linespread{1.66}
\setlength{\parindent}{0.5in}

\setcounter{secnumdepth}{0}

\pagestyle{empty}

\renewcommand{\section}[1]{%
\bigskip
\begin{center}
\begin{Large}
\normalfont\scshape #1
\medskip
\end{Large}
\end{center}}

\renewcommand{\subsection}[1]{%
\bigskip
\begin{center}
\begin{large}
\normalfont\itshape #1
\end{large}
\end{center}}

\renewcommand{\subsubsection}[1]{%
\vspace{2ex}
\noindent
\textit{#1.}---}

\renewcommand{\tableofcontents}{}

\bibpunct{(}{)}{;}{a}{}{,}  % this is a citation format command for natbib

@@

\title{How cryptic is cryptic diversity? Machine learning approaches to plastral variation in \textit{Emys marmorata}.}
\author[1]{Peter D Smits}%\thanks{psmits@uchicago.edu}}
\author[2]{Kenneth D Angielczyk}%\thanks{kangielczyk@fieldmuseum.org}}
\author[3]{James F Parham}%\thanks{jparham@fullerton.edu}}
\affil[1]{Committee on Evolution Biology, University of Chicago}
\affil[2]{Department of Geology, Field Museum of Natural History}
\affil[3]{Department of Geological Sciences, California State University -- Fullerton}

\begin{document}
\maketitle
\noindent{\textbf{Crresponding author:} Peter D Smits, Committee on Evolutionary Biology, University of Chicago, ADDRESS, STATE, ZIP, USA; E-mail: \href{mailto:psmits@uchicago.edu}{psmits@uchicago.edu}}

%\linenumbers
%\modulolinenumbers[2]

\begin{abstract}
  % 200 words
\noindent (Keywords: Testudines, morphology, geometric morphometrics)
\end{abstract}

%\section{Introduction}

% cryptic diversity
%   most species are still deliminated solely based on morphology
Cryptic diversity is when taxa were only first deliminated via molecular means and were not or cannot deliminated via morphological identification CITATION. The discovery of this previously unknown diversity has expanded diversity amoungst certain taxa CITATION. Additionally, a suite of methods have been developed to more accurately deliminate taxa based on molecular divergence CITATION.
% all the work done on species delimitation strictly from molecular data

%   paleo problem
Concerns about cryptic diversity are enhanced when taxa are only known from morphology, such as the case of extinct organisms CITATION.

@@

% GP superimposition and PCA
``Half'' plastra landmark configurations were superimposed using generalized Procrustes analysis \citep{Dryden1998a} after which, the principal components (PC) of shape were calculated. This was done using the \texttt{shapes} package for R \citep{2013, Dryden2013}.

@@

% model training
%   tuning parameters
%     grid search
%       what are the tuning parameters for each model?
%     best AUC ROC
For both types of supervised learning models, tuning parameters were estimated using 10 rounds of 10-fold cross-validation (CV) across a grid search of all tuning parameter combinations. Optimal tuning parameter values were selected based on area under the receiver operating characteristic curve (AUC ROC). Multiclass AUC ROC was estimated using the all-against one strategy derived by \citet{Hand2001} in implemented in PROC PACKAGE.

@@

\subsubsection{Supervised learning}
%For all of the classification schemes, the optimal models for both the multinomial logistic regression (Table SUPPLEMENT) and random forest (Fig. \ref{fig:roc}) methods include many, if not most, of the possible features.
<<>>=
daic <- unlist(lapply(lapply(tm.a.analysis$sel, function(x) x$delta), 
                      function(x) x[2]))
@
% multinomial logistic regression model
The optimal multinomial logistic regression models for all four classification schemes, based on AICc, have all \Sexpr{length(tm.a.analysis$best$spinks$coefnames)} possible PCs as features(Table SUPPLEMENT). The second best models for all classification schemes have \Sexpr{length(tm.a.analysis$best$spinks$coefnames) - 1} PCs as features. The \(\delta\)AICc values between the optimal and second best model range from \Sexpr{min(daic)} for the first morphological based classification hypothesis to \Sexpr{max(daic)} for the second molecular based classification hypothesis.

While the \(\delta\)AICc value between the optimal and second best model for the first morphological based classification hypothesis is within the range to be considered sufficient and equally optimal \citep{Burnham2002a}, for this analysis we chose to use only the absolute AICc best model. While AICc values can not be compared between models with different responses \citep{Burnham2002a}, we interpret the fact that the optimal model for all classification schemes is the global model as a reason to just use this model structure for all cases. Additionally, by maintaining a single optimal model for all classification schemes, this limits the total number of comparisons between the bootstrap resampled distributions of the AUC ROC values for the testing data set (see below).

% random forest model
<<>>=
oprf <- unlist(lapply(trf.a, function(x) x$optsize))
@

The selected number of features in the optimal random forest models for each classification scheme vary much more than in the case of the multinomial logistic regression models (Fig. \ref{fig:roc}), ranging from \Sexpr{min(oprf)} for the first morphological based classification hypothesis to \Sexpr{max(oprf)} for the second morphological based classification hypothesis.

% need to discuss the final models in both cases.
% figure
%   ROC model selection (see talk)
%   generalize densities
%     facet: multinomial, rf (see talk)

In the case of all models, there is a substantial increase in model performance as measured by AICc for the multinomial logistic models (Tables SUPPLEMENT) or in AUC ROC for the random forest models and illustrated for the multinomial logistic regression models as the number of features increases (Fig. \ref{fig:roc}).

@@

This pattern of misclassification may be caused by the differences in mean shape between each of the different classes (Fig. \ref{fig:mean_shape}). The mean shape of the northern clade is the most similar to the mean shape of the entire dataset (Fig. \ref{fig:mean_shape}\ref{fig:imp_var}), which may mean that specimens that are closer to the mean shape will be systematically misclassified as the northern clade.

@@

% methodological concerns
%   compromise in the supervised learning models
%     fair because as much variation as ``necessary'' is used
The methods presented here for supervised learning analysis of the landmark variation represent a compromise

%   unsupervised learning
%     model based clustering
%     nonparametric bayesian approach
%     anything about semisupervised clustering? probably not
Ultimately, it would be optimal to not require such explicit classification hypotheses, especially when concerned about possible cryptic variation in extinct taxa. The method employed in this study, PAM, is rather simple and not model based. Comparison is facilitated by comparison of a summary statistic and bootstrap confidence intervals. A more useful and future avenue would be employing various model based clustering approaches CITATIONS. In this manner, a series of candidate models can be compared via model comparison methods, such as AIC or Bayes factors CITATION, in order to asses the best clustering solution. Of particular note are nonparametric Bayesian approaches to model based clustering CITATIONS. This approach uses a class of flexible priors to allow for the most optimal clustering solution to be decided from the data. Currently, there exists a nonparametric Bayesian clustering method and further development of this approach may prove extremely fruitful for better deliminating taxa solely from morphology.

@@

\section{Acknowledgements}
PDS would like to thank David Bapst, Michael Foote, Benjamin Frable, and Dallas Krentzel for useful discussion which enhanced the quality of this study.

### 2013-07-15 documents/turtle_paper.tex
more draft work

\documentclass[12pt]{article}\usepackage{graphicx, color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

@@

\usepackage{alltt}
\usepackage{amsmath, amsthm}
\usepackage{graphicx, microtype, hyperref, authblk}
\usepackage{rotating, longtable, caption, subcaption, multirow}
\usepackage[sort&compress]{natbib}
%\usepackage{fullpage}

@@

\title{How cryptic is cryptic diversity? Machine learning approaches to plastral variation in \textit{Emys marmorata}.}
\author[1]{Peter D Smits \thanks{psmits@uchicago.edu}}
\author[2]{Kenneth D Angielczyk \thanks{kangielczyk@fieldmuseum.org}}
\author[3]{James F Parham \thanks{jparham@fullerton.edu}}
\affil[1]{Committee on Evolution Biology, University of Chicago}
\affil[2]{Department of Geology, Field Museum of Natural History}
\affil[3]{Department of Geological Sciences, California State University -- Fullerton}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}

\begin{document}

\maketitle

\linenumbers
\modulolinenumbers[2]

\begin{abstract}
  % 200 words

\end{abstract}

\section{Introduction}

% cryptic diversity
%   most species are still deliminated solely based on morphology
%   paleo problem
Cryptic diversity is when taxa were only first deliminated via molecular means and were not or cannot deliminated via morphological identification CITATION. The discovery of this previously unknown diversity has
% all the work done on species delimitation strictly from molecular data
%

@@

\subsubsection{Supervised learning}
For all classification schemes, the optimal random forest model based on recursive feature selection by maximizing AUC ROC of the model was one with many features (Fig. \ref{fig:roc})
% need to discuss the final models in both cases.
% figure
%   ROC model selection (see talk)
%   generalize densities
%     facet: multinomial, rf (see talk)

>>>>

\documentclass[12pt,letterpaper]{article}\usepackage{graphicx, color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

@@

\usepackage{alltt}
\usepackage{amsmath, amsthm}
\usepackage{graphicx, microtype, hyperref, authblk}
\usepackage{rotating, longtable, caption, subcaption, multirow}
\usepackage[sort&compress]{natbib}
\usepackage{fullpage}

@@

\title{How cryptic is cryptic diversity? Machine learning approaches to plastral variation in \textit{Emys marmorata}.}
\author[1]{Peter D Smits}%\thanks{psmits@uchicago.edu}}
\author[2]{Kenneth D Angielczyk}%\thanks{kangielczyk@fieldmuseum.org}}
\author[3]{James F Parham}%\thanks{jparham@fullerton.edu}}
\affil[1]{Committee on Evolution Biology, University of Chicago}
\affil[2]{Department of Geology, Field Museum of Natural History}
\affil[3]{Department of Geological Sciences, California State University -- Fullerton}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}

\begin{document}
\maketitle
\noindent{\textbf{Crresponding author:} Peter D Smits, Committee on Evolutionary Biology, University of Chicago, ADDRESS, STATE, ZIP, USA; E-mail: \href{mailto:psmits@uchicago.edu}{psmits@uchicago.edu}}

%\linenumbers
%\modulolinenumbers[2]

\begin{abstract}
  % 200 words
\noindent (Keywords: Testudines, morphology, geometric morphometrics)
\end{abstract}

%\section{Introduction}

% cryptic diversity
%   most species are still deliminated solely based on morphology
Cryptic diversity is when taxa were only first deliminated via molecular means and were not or cannot deliminated via morphological identification CITATION. The discovery of this previously unknown diversity has expanded diversity amoungst certain taxa CITATION. Additionally, a suite of methods have been developed to more accurately deliminate taxa based on molecular divergence CITATION.
% all the work done on species delimitation strictly from molecular data

%   paleo problem
Concerns about cryptic diversity are enhanced when taxa are only known from morphology, such as the case of extinct organisms CITATION.

@@

\subsubsection{Supervised learning}
%For all of the classification schemes, the optimal models for both the multinomial logistic regression (Table SUPPLEMENT) and random forest (Fig. \ref{fig:roc}) methods include many, if not most, of the possible features.

% multinomial logistic regression model
The optimal multinomial logistic regression models for all four classification schemes, based on AICc, have all 10 possible PCs as features(Table SUPPLEMENT). The second best models for all classification schemes have 9 PCs as features. The \(\delta\)AICc values between the optimal and second best model range from 2.0639 for the first morphological based classification hypothesis to 19.8349 for the second molecular based classification hypothesis.

While the \(\delta\)AICc value between the optimal and second best model for the first morphological based classification hypothesis is within the range to be considered sufficient and equally optimal \citep{Burnham2002a}, for this analysis we chose to use only the absolute AICc best model. While AICc values can not be compared between models with different responses \citep{Burnham2002a}, we interpret the fact that the optimal model for all classification schemes is the global model as a reason to just use this model structure for all cases. Additionally, by maintaining a single optimal model for all classification schemes, this limits the total number of comparisons between the bootstrap resampled distributions of the AUC ROC values for the testing data set (see below).

% random forest model

The selected number of features in the optimal random forest models for each classification scheme vary much more than in the case of the multinomial logistic regression models (Fig. \ref{fig:roc}), ranging from 6 for the first morphological based classification hypothesis to 10 for the second morphological based classification hypothesis.

% need to discuss the final models in both cases.
% figure
%   ROC model selection (see talk)
%   generalize densities
%     facet: multinomial, rf (see talk)

In the case of all models, there is a substantial increase in model performance as measured by AICc for the multinomial logistic models (Tables SUPPLEMENT) or in AUC ROC for the random forest models and illustrated for the multinomial logistic regression models as the number of features increases (Fig. \ref{fig:roc}).
